The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2017-01-12 script/de/chapters/numbertheory.tex
CT Book ed 12: Small text correction in chapter 4.19.4 concerning the Sage samples 4.15 and 4.16 about primroots.

Die Datei \glqq primroots\_1-100000.dat\grqq~enthält die Liste aller Primitivwurzeln
% Either mask _ within brackets, or use \verb!!primroots_1-100000.dat!
für alle Primzahlen zwischen 1 und 100.000 inklusive. Dies ist eine große Datei
(ca. 1,1 GB unkomprimiert, und 156 MB komprimiert mit 7Zip).
Die komprimierte Datei kann abgerufen werden unter\\
\url{https://www.cryptool.org/images/ctp/documents/primroots_1-100000.7z}.

Im Bereich 1 bis 100.000 liegen 9.592 Primzahlen. Für sie wurden insgesamt mehr als 
169 Millionen Primitivwurzeln berechnet. Für jede Primzahl p>3 gilt, dass zwischen
20 \% und knapp 50 \% aller Zahlen zwischen 1 und p eine Primitivwurzel davon darstellen
und dass die Anzahl der Primitivwurzeln immer eine gerade Zahl ist.
% Oder anders:
% Theorem: If p is a prime, then there is a primitive root for p.
% (In fact, if p is a prime and p>2, there are always at least two primitive roots,
% and sometimes many more.)
% The proof of this theorem is hard. It's in our ??? text (9.2?) and in
% second-semeter abstract algebra.
Eine Formel, die die Anzahl der Primitivwurzeln zu einer Zahl angibt, kennen wir nicht.

Wenn man die kleinste Primitivwurzel sucht ... diese wurden für alle Primzahlen
bis zu einer Million mit dem Sage-Skript \ref{l:Sagecode_SmallestPrimRoot} berechnet
und in der Datei \glqq primroot-smallest\_up-to-one-million.dat\grqq~gespeichert.
%%%   : ca. xxxxxxxxx,1 GB unkomprimiert, und xxxxx MB komprimiert mit 7Zip.
Die komprimierte Datei kann abgerufen werden unter
\url{https://www.cryptool.org/images/ctp/documents/primroot-smallest_up-to-one-million.7z}.

Die Ergebnis-Datenbank enthält für jede einzelne Primzahl p zwischen 1 und 1.000.000
die Anzahl der zugehörigen Primitivwurzeln und die kleinste Primitivwurzel mod p.
Dies kann für Zahlentheoretiker interessant sein.\footnote{%
  Siehe auch Re: [sage-devel] What can we do with a database of primitive roots?
  \url{https://groups.google.com/forum/m/#!topic/sage-devel/TA5Nk2GdhOI}
}

% David Joyner mentions in his answer in [sage-devel]:
% "Least primitive root of prime numbers"
% http://plouffe.fr/simon/articles/Least%20primitive%20root%20of%20prime%20numbers.pdf
% Tomás Oliveira e Silva  October 21, 2004 (Universidade de Aveiro, PORTUGAL)

% Vgl. später auch mal:
% - http://wstein.org/edu/124/lectures/lecture11/lecture11.tex
% - http://www.aw-bc.com/rosen/weblinkch9.html   Annotated Web Links to
%   CHAPTER 9 - Primitive Roots in Elementary Number Theory, 5/E by
%   Kenneth H. Rosen, AT&T Laboratories, Publisher: Addison-Wesley, 2005, 744 pages
% - http://arxiv.org/pdf/0810.2325.pdf
%   A NOTE ON ARTIN’S CONSTANT  by  IVAN CHEREDNIK
% xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

\newpage
Das Sage-Skript \ref{l:Sagecode_SmallestPrimRoot} berechnet alle Primitivwurzeln
für alle Primzahlen bis zu einer Million, und gibt jeweils die Anzahl unterschiedlicher
Primitivwurzeln und die jeweils kleinste Primitivwurzel aus.

@@

\vskip +20 pt
Wenn man pro Primzahl nur die kleinste Primitivwurzel sucht, könnte man dieses Skript
drastisch beschleunigen, indem man anhand der Theorie die möglichen Kandidaten direkter
sucht (anstatt mit \verb!enum_PrimitiveRoots_of_an_Integer! erstmal alle Primroots zu
generieren).

>>>>

Im Bereich 1 bis 100.000 liegen 9.592 Primzahlen. Für sie wurden insgesamt mehr als 
169 Millionen Primitivwurzeln berechnet. Für jede Primzahl $p>3$ gilt, dass zwischen
20 \% und knapp 50 \% aller Zahlen zwischen 1 und p eine Primitivwurzel davon darstellen.
%%%%% Oder anders:
% Theorem: If p is a prime, then there is a primitive root for p.
% (In fact, if p is a prime and p>2, there are always at least two primitive roots,
% and sometimes many more.)
% The proof of this theorem is hard. It's in our ??? text (9.2?) and in
% second-semeter abstract algebra.

Die resultierende Datei \glqq primroots\_1-100000.dat\grqq~enthält die Liste
% Either mask _ within brackets, or use \verb!!primroots_1-100000.dat!
aller Primitivwurzeln für alle Primzahlen zwischen 1 und 100.000 inklusive.
Dies ist eine große Datei (ca. 1,1 GB unkomprimiert, und 156 MB komprimiert mit 7Zip).
Die komprimierte Datei kann abgerufen werden unter\\
\url{https://www.cryptool.org/images/ctp/documents/primroots_1-100000.7z}.

Ihr Inhalt sieht wie folgt aus:
\begin{Verbatim}%
[fontsize=\footnotesize]
2      1      {1}
3      1      {2}
5      2      {2, 3}
7      2      {3, 5}
11     4      {8, 2, 6, 7}
...
99989  42840  {2, 3, 8, 10, 11, 13, 14, ..., 99978, 99979, 99981, 99986, 99987}
99991  24000  {65539, 6, 65546, 11, 12, ..., 65518, 65520, 87379, 65526, 65528}
\end{Verbatim}

\newpage
Das Sage-Skript \ref{l:Sagecode_SmallestPrimRoot} berechnet für jede Primzahl (bis \glqq end\grqq)
alle Primitivwurzeln, und gibt jeweils die Anzahl unterschiedlicher
Primitivwurzeln und die jeweils kleinste Primitivwurzel aus.

@@

Das Sage-Skript \ref{l:Sagecode_SmallestPrimRoot} wurde nach mehreren Wochen
gestoppt (wo es auf einem modernen PC mit SageMath 7.2 lief), nachdem es alle
Primzahlen bis zu einer halben Million auswertete. Das Ergebnis wurde in der
Datei \glqq primroot\_number-of-and-smallest\_up-to-prime-500107.dat\grqq~gespeichert
(sie ist unkomprimiert 617 kB groß, und 178 kB komprimiert mit 7Zip).

Die komprimierte Datei kann abgerufen werden unter
%%% \url{https://www.cryptool.org/images/ctp/documents/primroot-smallest_up-to-one-million.7z}.
\url{https://www.cryptool.org/images/ctp/documents/primroot_number-of-and-smallest_up-to-prime-500107.7z}.

Diese Datenbank enthält für jede einzelne Primzahl p zwischen 1 und 500.107
die Anzahl der zugehörigen Primitivwurzeln und die kleinste Primitivwurzel mod p.
Es gilt, dass die Anzahl der Primitivwurzeln (für $p>3$) immer gerade ist.
Eine Formel, die die Anzahl der Primitivwurzeln zu einer Zahl angibt, kennen wir nicht.
Diese Datenbank könnte für Zahlentheoretiker interessant sein.\footnote{%
  Siehe auch Re: [sage-devel] What can we do with a database of primitive roots?\\
  \url{https://groups.google.com/forum/m/#!topic/sage-devel/TA5Nk2GdhOI}
}
% David Joyner mentions in his answer in [sage-devel]:
% "Least primitive root of prime numbers"
% http://plouffe.fr/simon/articles/Least%20primitive%20root%20of%20prime%20numbers.pdf
% Tomás Oliveira e Silva  October 21, 2004 (Universidade de Aveiro, PORTUGAL)
%
% Vgl. später auch mal:
% - http://wstein.org/edu/124/lectures/lecture11/lecture11.tex
% - http://www.aw-bc.com/rosen/weblinkch9.html   Annotated Web Links to
%   CHAPTER 9 - Primitive Roots in Elementary Number Theory, 5/E by
%   Kenneth H. Rosen, AT&T Laboratories, Publisher: Addison-Wesley, 2005, 744 pages
% - http://arxiv.org/pdf/0810.2325.pdf
%   A NOTE ON ARTIN’S CONSTANT  by  IVAN CHEREDNIK
% xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx
%
Ihr Inhalt sieht wie folgt aus:
\begin{Verbatim}%xxxxxxxxxxx
[fontsize=\footnotesize]
2       1       1
3       1       2
5       2       2
7       2       3
11      4       2
13      4       2
17      8       3
...
99989   42840   2
99991   24000   6
100003  28560   2
...
500069  250032  2
500083  151520  2
500107  156864  2
\end{Verbatim}

\vskip +20 pt
Wenn man pro Primzahl nur die kleinste Primitivwurzel sucht, kann man dieses Skript
drastisch beschleunigen, indem man anhand der Theorie die möglichen Kandidaten direkter
sucht (anstatt mit \verb!enum_PrimitiveRoots_of_an_Integer! erstmal alle Primroots zu
generieren).

### 2017-01-12 script/en/chapters/numbertheory.tex
CT Book ed 12: Small text correction in chapter 4.19.4 concerning the Sage samples 4.15 and 4.16 about primroots.

\newpage
\noindent A variant of our function \verb!count_PrimitiveRoots_of_a_PrimesRange!
slightly modified by Minh Van Nguyen was used to generate a database of all
primitive roots of all primes between 1 and 100,000.

@@

The resulting file ``primroots\_1-100000.dat'' is a database of all primitive roots of all
% Either mask _ within brackets, or use \verb!!primroots_1-100000.dat!
primes between 1 and 100,000 inclusive. It is a large file (about 1.1 GB
uncompressed, and 156 MB compressed with 7Zip).
You can find the compressed file at
\url{https://www.cryptool.org/images/ctp/documents/primroots_1-100000.7z}.

Between 1 and  100,000 there are 9,592 primes. For them, more than 169 million primitive
roots have been calculated. For each prime p>3 it holds that between 20 \% and almost
50 \% of all integers between 1 and p are an according primitive root and that the
number of primitive roots is always an odd number. We don't know a formula which delivers
the number of primitive roots for a given number.

If you are looking for the smallest primitive root ... they have been calculated
for all primes out to a million with the Sage script \ref{l:Sagecode_SmallestPrimRoot}
and stored in  file ``primroot-smallest\_up-to-one-million.dat''.
%%%   : about xxxxx.1 GB uncompressed, and xxxxx MB compressed with 7Zip.
You can find the compressed file at
\url{https://www.cryptool.org/images/ctp/documents/primroot-smallest_up-to-one-million.7z}.

The resulting database contains all primes p between 1 and 1,000,000 together with the
according number of primitive roots and the according smallest prim root mod p. This
may be interesting to some number theorists.\footnote{%
  See Re: [sage-devel] What can we do with a database of primitive roots?
  \url{https://groups.google.com/forum/m/#!topic/sage-devel/TA5Nk2GdhOI}
}

@@

>>>>

\newpage
\noindent A variant of our function \verb!count_PrimitiveRoots_of_a_PrimesRange!
(slightly modified by Minh Van Nguyen) was used to generate a database of all
primitive roots of all primes between 1 and 100,000.

@@

Between 1 and 100,000 there are 9,592 primes. For them, more than 169 million primitive
roots have been calculated. For each prime $p>3$ it holds that between 20 \% and almost
50 \% of all integers between 1 and p are an according primitive root.

The resulting file ``primroots\_1-100000.dat'' is a database of all primitive roots of all
% Either mask _ within brackets, or use \verb!!primroots_1-100000.dat!
primes between 1 and 100,000 inclusive. It is a large file (about 1.1 GB
uncompressed, and 156 MB compressed with 7Zip).
You can find the compressed file at\\
\url{https://www.cryptool.org/images/ctp/documents/primroots_1-100000.7z}.

Its content looks like this:
\begin{Verbatim}%
[fontsize=\footnotesize]
2      1      {1}
3      1      {2}
5      2      {2, 3}
7      2      {3, 5}
11     4      {8, 2, 6, 7}
...
99989  42840  {2, 3, 8, 10, 11, 13, 14, ..., 99978, 99979, 99981, 99986, 99987}
99991  24000  {65539, 6, 65546, 11, 12, ..., 65518, 65520, 87379, 65526, 65528}
\end{Verbatim}

@@

%%% \vskip +20 pt
The Sage script \ref{l:Sagecode_SmallestPrimRoot} was stopped after several weeks
(running on a modern PC with SageMath 7.2) after investigating all primes up to
half a million. The result was stored in the file
``primroot\_number-of-and-smallest\_up-to-prime-500107.dat'', which is 617 kB
uncompressed, and 178 kB compressed with 7Zip.

You can find the compressed file at
%%% \url{https://www.cryptool.org/images/ctp/documents/primroot-smallest_up-to-one-million.7z}.
\url{https://www.cryptool.org/images/ctp/documents/primroot_number-of-and-smallest_up-to-prime-500107.7z}.

This database contains all primes p between 1 and 500,107 together with the
according number of primitive roots and the according smallest prim root mod p.
It holds that the number of primitive roots (for $p>3$ )is always an odd number.
We don't know a formula which delivers the number of primitive roots for a given number.
So this database may be interesting to some number theorists.\footnote{%
  See Re: [sage-devel] What can we do with a database of primitive roots?\\
  \url{https://groups.google.com/forum/m/#!topic/sage-devel/TA5Nk2GdhOI}
}
Its content looks like this:
\begin{Verbatim}%xxxxxxxxxxx
[fontsize=\footnotesize]
2       1       1
3       1       2
5       2       2
7       2       3
11      4       2
13      4       2
17      8       3
...
99989   42840   2
99991   24000   6
100003  28560   2
...
500069  250032  2
500083  151520  2
500107  156864  2
\end{Verbatim}
